Climate variability indices for ecological and crop models in R: the `climatrends` package

### 2020-05-22 inst/paper/paper.Rmd
speed-gdd-late_frost

---
title: 'climatrends: Precipitation and temperature indices for climate variability analysis in R'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
output: pdf_document
---

@@

\
```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("magrittr")
funtbl <- read.csv("available_functions.csv")
```

@@

Six main functions are provided, `crop_sensitive()`, `ETo()`, `GDD()`, `late_frost()`, `rainfall()` and `temperature()` with a default method for numeric 'vector's and additional methods [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The later two are designed to fetch data from cloud sources.

@@

@@

Additionally, the function `GDD()` offers three modified equations (two) designed for cold environments and (one) for tropical environments. For cold environments, where $T_{min}$ may be lower than $T_{0}$, there are two variants of the previous equation to adjust either $T_{mean}$ (variant a) or $T_{min}$ (variant b). The variant a changes $T_{mean}$ to $T_{0}$ if $T_{mean} < T_{0}$ and is expressed as follows.

@@

By default, the function returns the degree-days that is accumulated over the time series using Equation 1. Additionally, the function may return the daily values of degree-days or the number of days that a given organism required to reach a certain number of accumulated degree-days. These values are defined by 'acc', 'daily' or 'ndays' and can be adjusted using the argument `return.as`. The required accumulated gdd is defined with argument `degree.days`. For example, the Korean pine (*Pinus koraiensis*) requires 105 $^\circ C$ accumulated gdd to onset the photosynthesis [@JWu2013]. In that case, `GDD()` will calculate the growing degree-days ($gdd$) and sum up the values until it reaches the defined gdd (105 $^\circ C$) and return the number of days needed in the given season ($GDD_{ndays}$), as follows.

@@

$$\parallel GDD_{ndays} \parallel \: = \; ggd_1 \;+ \; ...  \; +  \; gdd_n$$

## Reference evapotranspiration

@@

@@

The percentage of daytime hours ($p$) is calculated internally by most methods in `ETo()` (except the array method) using the given latitude (taken from the inputted `object`) and date (taken from the inputted `day.one`). It matches the latitude and date with a table of daylight percentage derived from Brouwer and Heibloem [-@Brouwer1986]. The table can be verified using `climatrends:::daylight`.

@@

>>>>

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
output: pdf_document
---

```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("magrittr")
funtbl <- read.csv("available_functions.csv")
```

@@

\

@@

Six main functions are provided, `crop_sensitive()`, `ETo()`, `GDD()`, `late_frost()`, `rainfall()` and `temperature()` with a default method for numeric 'vector's and additional methods implemented via the package `methods` [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The later two are designed to fetch data from cloud sources.

@@

## Temperature and precipitation indices

@@

Additionally, the function `GDD()` offers three modified equations designed for cold environments and for tropical environments. For cold environments, where $T_{min}$ may be lower than $T_{0}$, there are two modified equations that adjusts either $T_{mean}$ (variant a) or $T_{min}$ (variant b). The variant a changes $T_{mean}$ to $T_{0}$ if $T_{mean} < T_{0}$ and is expressed as follows.

@@

By default, the function returns the degree-days that is accumulated over the time series using Equation 1. Additionally, the function may return the daily values of degree-days or the number of days that a given organism required to reach a certain number of accumulated degree-days. These values are defined by 'acc', 'daily' or 'ndays' and can be adjusted using the argument `return.as`. The required accumulated gdd is defined with argument `degree.days`. For example, the Korean pine (*Pinus koraiensis*) requires 105 $^\circ C$ accumulated gdd to onset the photosynthesis [@JWu2013]. In that case, `GDD()` will calculate the growing degree-days ($gdd$) and sum up the values until it reaches 105 $^\circ C$ and return the number of days required in the given season ($GDD_{r}$), as follows.

@@

$$\parallel GDD_{r} \parallel \: = \; ggd_1 \;+ \; ...  \; +  \; gdd_n$$

## Late-spring frost

## Crop-related indices

Two functions in **climatrends** are mainly designed to capture the effects of climate on the development and stress of crop species, `crop_sensitive` computes indices that aims to capture the changes in temperature extremes during key phenological stages (e.g. anthesis), and `ETo()` computes the reference evapotranspiration.

@@

Equation [6]

@@

The percentage of daytime hours ($p$) is calculated internally by the 'data.frame' and 'sf' methods in `ETo()` using the given latitude (taken from the inputted `object`) and date (taken from the inputted `day.one`). It matches the latitude and date with a table of daylight percentage derived from Brouwer and Heibloem [-@Brouwer1986]. The table can be verified using `climatrends:::daylight`.

@@

# Authors' contributions

KdS and JvE conceived the ideas of implementing an `R` package. JvE and SØS assisted in the implementation of methods. KdS wrote and documented the software. [add other contributions...]

# Data availability statement

The source code has been archived at [**add link**] as `climatrends` version **X.Y.Z**. To explore the latest functionalities of `climatrends`, please check the package's updates at CRAN (https://cran.r-project.org/package=climatrends).
